Inverse conformational selection in lipid-protein binding

### 2021-05-28 manuscriptPGPE.tex
Update text

@@

Chemical compositions of hydrophilic lipid headgroups vary between different
organelles and organisms, and different lipid types
regulate protein functions in many different ways \cite{lee03,vanmeer08}.
Lipids can directly bind to proteins, or indirectly affect protein
functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
Specific interactions with certain lipid headgroups
are known to be essential for the function of several proteins \cite{lee11,lemmon08},
but it is not clear if the specificity arises from the
differences in accessible conformations between lipid types or
from specific intermolecular lipid--protein interactions.

@@

Here, we use natural abundance $^1$H-$^{13}$C solid-state NMR experiments and MD simulations from the NMRlipids Project
to resolve differences in the conformational ensembles of PC, PE, PG, and PS lipid headgroups.
Zwitterionic PC are the most common lipids in eukaryotes and PE in bacteria \cite{vanmeer08,sohlenkamp16};
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS are the most common negatively charged lipids in eukaryotes and PG in bacteria,
and both are known to affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
All the four studied lipid types specifically bind to various proteins \cite{yeagle14};
we therefore elucidate also the effect of protein binding to headgroup conformations.
These interactions are crucial, for example, in lipid-mediated signaling \cite{lemmon08} and for the
design of phoshopolipid-specific antibodies \cite{vigant15}.

@@

Our results pave the way toward understanding lipid-mediated cell signaling and how lipids regulate membrane protein function in general.
We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids. The broad conformational ensembles we see in bulk bilayers suggest that the structures observed in pure lipid crystals occur seldomly in liquid state. Consequently, in protein binding the entropic cost on the lipid may be significant.

Finally, our work has demonstrated the power of open access MD simulation data
to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.

>>>>

\begin{tocentry}

     \includegraphics[height=3.5cm]{../Figs/TOC2.eps}
%Some journals require a graphical entry for the Table of Contents.
%This should be laid out ``print ready'' so that the sizing of the
%text is correct.

%Inside the \texttt{tocentry} environment, the font used is Helvetica
%8\,pt, as required by \emph{Journal of the American Chemical
%Society}.

%The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
%permitted for  \emph{Journal of the American Chemical Society}
%graphical table of content entries. The box will not resize if the
%content is too big: instead it will overflow the edge of the box.

%This box and the associated title will always be printed on a
%separate page at the end of the document.

\end{tocentry}

@@

Lipid interactions with other biomolecules is gaining interest
not only in molecular cell biology \cite{harayama18}, but also in
nanobiotechnology applications, such as design of antibodies \cite{vigant15}
and carriers for mRNA vaccines \cite{pardi18,schoenmaker21}.
Interactions of cellular membranes or lipid nanoparticles with other biomolecules,
such as proteins, drugs or RNA, are regulated by lipid headgroups,
the outward facing components of lipid bilayers \cite{harayama18,vanmeer08}.
%Chemical compositions of hydrophilic  vary between different
%organelles and organisms, and different lipid types
%regulate protein functions in many different ways \cite{lee03,vanmeer08}.
%Lipids can directly bind to proteins, or indirectly affect protein
%functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
Chemical compositions of lipid headgroups vary between different
organelles and organisms, and specific interactions with certain lipid headgroups
are known to be essential for the function of several proteins \cite{harayama18,vanmeer08}.
However, the driving forces for specific interactions between lipids and membrane binding
substances are not fully understood because conformational ensembles of lipids in
physiologically relevant liquid state have not been experimentally determined.
Therefore, it is not clear if the specificity arises from the
differences in accessible conformations between lipid types or
from specific intermolecular lipid--protein interactions.

@@

Here, we use natural abundance $^1$H-$^{13}$C solid-state NMR experiments and MD simulations from the NMRlipids Project
to resolve differences in the conformational ensembles of PC, PE, PG, and PS lipid headgroups.
Zwitterionic PC are the most common lipids in eukaryotes and PE in bacteria \cite{vanmeer08,sohlenkamp16};
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PC lipids are also used in mRNA COVID-19 vaccines \cite{schoenmaker21}.
PS are the most common negatively charged lipids in eukaryotes and PG in bacteria,
and both are known to affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
All the four studied lipid types specifically bind to various proteins \cite{yeagle14};
we therefore elucidate also the effect of protein binding to headgroup conformations.
%These interactions are crucial, for example, in lipid-mediated signaling \cite{lemmon08} and for the
%design of phoshopolipid-specific antibodies \cite{vigant15}. 
The resulting {\it inverse conformational selection model} for lipid-protein binding can be also
applied to understand interactions between membranes and other biomolecules than proteins, such as drugs, RNA, or viruses.
On the other hand, the detailed understanding of lipid bilayer interfaces can facilitate the design
of lipid nanoparticle carriers for mRNA vaccines with less side effects.

@@

Our results suggest that lipid binding to proteins can be described by the {\it inverse conformational selection model}
where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
In this model, the binding energy is dominated by
intermolecular interactions and changes in lipid conformational entropy, while 
conformational restrictions inherent to the lipids play a minor role.
%We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids. The broad conformational ensembles we see in bulk bilayers suggest that the structures observed in pure lipid crystals occur seldomly in liquid state. Consequently, in protein binding the entropic cost on the lipid may be significant.
Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
and functions of lipid regulated membrane proteins in general. 
In bionanotechnology, our model facilitates, for example,
the design of phoshopolipid-specific antibodies and lipid nanoparticle carries with minimal sides effects for mRNA vaccines.
Furthermore, our work has demonstrated the power of open access MD simulation data
to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.
